---
title: "data manipulation with 'dplyr'"
output: github_document
date: "2022-09-22"
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)

options(tibble.print_min = 3)

litters_data = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data = janitor::clean_names(litters_data)

pups_data = read_csv("./data/FAS_pups.csv",
  col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data)

```


## select

can select a subset of data, specify which columns you want to keep
```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive)
```

select a range of columns to include:
```{r}
select(litters_data, group:gd_of_birth)
```

specify R which column you don't want to include, can delete multiple variables in one line:
```{r}
select(litters_data, -pups_survive)
```

rename the variable:
```{r}
select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

also can use "rename" feature:
```{r}
rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

if variables are named with suffixes or other standard patterns and want to select those variables:
```{r}
select(litters_data, starts_with("gd"), ends_with("weight"))

```

to reorganize columns:
```{r}
select(litters_data, litter_number, pups_survive, everything())
```

also can use "relocate":
```{r}
relocate(litters_data, litter_number, pups_survive)
```

## filter

some filter options:
gd_of_birth == 20
pups_born_alive >= 2
pups_survive != 4
!(pups_survive == 4)
group %in% c("Con7", "Con8")
group == "Con7" & gd_of_birth == 20

```{r}
filter(litters_data, group == "Con7" & gd_of_birth == 20)
```


also can use `drop`:

remove any row with a missing value:
```{r}
drop_na(litters_data)
```

remove rows for which wt_increase is missing (see p8105 webpage):


## mutate

```{r}
mutate(litters_data,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
)
```


## arrange
```{r}
head(arrange(litters_data, group, pups_born_alive), 10)
```


## %>%

example 1:
```{r}
litters_data_raw = read_csv("./data/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data_clean_names = janitor::clean_names(litters_data_raw)
litters_data_selected_cols = select(litters_data_clean_names, -pups_survive)
litters_data_with_vars = 
  mutate(
    litters_data_selected_cols, 
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group))
litters_data_with_vars_without_missing = 
  drop_na(litters_data_with_vars, wt_gain)
litters_data_with_vars_without_missing
```

example 2:
```{r}
litters_data_clean = 
  drop_na(
    mutate(
      select(
        janitor::clean_names(
          read_csv("./data/FAS_litters.csv", col_types = "ccddiiii")
          ), 
      -pups_survive
      ),
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)
    ),
  wt_gain
  )
```


better approach using `%>%`
```{r}
litters_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  select(-pups_survive) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% 
  drop_na(wt_gain)
```


using `.`:
```{r}
litters_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names(dat = .) %>%
  select(.data = ., -pups_survive) %>%
  mutate(.data = .,
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% 
  drop_na(data = ., wt_gain)
```

another example:
```{r}
litters_data %>%
  lm(wt_gain ~ pups_born_alive, data = .) %>%
  broom::tidy()
```





